---
title: "QBS121_final"
output: pdf_document
date: "2025-03-01"
---

# Data EDA
## Load the data
```{r,include=FALSE, warning=FALSE}
data = read.csv('data.csv')
```

## Visualization of variables
```{r,echo=FALSE, warning=FALSE}
# Custom theme for better visualization
library(ggplot2)
library(gridExtra)
library(RColorBrewer)
library(viridis)
library(patchwork)

# Create custom theme
my_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Custom color palette
survival_colors <- c("#3498db", "#e74c3c")
```

### Visualization of the target variable
```{r,echo=FALSE, warning=FALSE}
# Count of death events
p1 <- ggplot(data, aes(x = factor(DEATH_EVENT, labels = c("Survived", "Died")))) +
  geom_bar(fill = survival_colors, width = 0.6) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Distribution of Outcomes",
       x = "Outcome",
       y = "Count") +
  my_theme

# Calculate percentages for pie chart
outcome_counts <- table(data$DEATH_EVENT)
outcome_pct <- round(100 * outcome_counts / sum(outcome_counts), 1)
pie_data <- data.frame(
  outcome = c("Survived", "Died"),
  count = as.numeric(outcome_counts),
  pct = as.numeric(outcome_pct)
)

# Pie chart
p2 <- ggplot(pie_data, aes(x = "", y = count, fill = outcome)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = survival_colors) +
  labs(title = "Percentage of Outcomes",
       fill = "Outcome") +
  geom_text(aes(label = paste0(pct, "%")), 
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 12, face = "bold", hjust = 0.5))

# Combine plots
p1 + p2 + plot_layout(ncol = 2)
```

### Explore numerical variables
```{r,echo=FALSE, warning=FALSE}
# Create a function for histogram + density plots
hist_density_plot <- function(variable, title, color, binwidth = NULL) {
  p <- ggplot(data, aes(x = !!sym(variable))) +
    geom_histogram(aes(y = after_stat(density)), 
                  fill = color, color = "white", alpha = 0.7,
                  binwidth = binwidth) +
    geom_density(alpha = 0.2, fill = color) +
    labs(title = title,
         x = variable,
         y = "Density") +
    my_theme
  return(p)
}

# Function for boxplots by outcome
boxplot_by_outcome <- function(variable, title, y_label) {
  p <- ggplot(data, aes(x = factor(DEATH_EVENT, labels = c("Survived", "Died")), 
                       y = !!sym(variable),
                       fill = factor(DEATH_EVENT, labels = c("Survived", "Died")))) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.fill = "white") +
    scale_fill_manual(values = survival_colors) +
    labs(title = title,
         x = "Outcome",
         y = y_label,
         fill = "Outcome") +
    my_theme +
    theme(legend.position = "none")
  return(p)
}

# Age plots
p1 <- hist_density_plot("age", "Age Distribution", "#3498db")
p2 <- boxplot_by_outcome("age", "Age by Outcome", "Age (years)")

# Ejection fraction plots
p3 <- hist_density_plot("ejection_fraction", "Ejection Fraction", "#2ecc71")
p4 <- boxplot_by_outcome("ejection_fraction", "Ejection Fraction by Outcome", "Ejection Fraction (%)")

# Serum creatinine plots
p5 <- hist_density_plot("serum_creatinine", "Serum Creatinine", "#9b59b6")
p6 <- boxplot_by_outcome("serum_creatinine", "Serum Creatinine by Outcome", "Serum Creatinine (mg/dL)")

# Combine plots
(p1 + p2) / (p3 + p4) / (p5 + p6)
```

```{r,echo=FALSE, warning=FALSE}
# Platelets plots
p1 <- hist_density_plot("platelets", "Platelets Distribution", "#f1c40f")
p2 <- boxplot_by_outcome("platelets", "Platelets by Outcome", "Platelets (kiloplatelets/mL)")

# Creatinine phosphokinase plots
p3 <- hist_density_plot("creatinine_phosphokinase", "CPK Distribution", "#e67e22", binwidth = 100)
p4 <- boxplot_by_outcome("creatinine_phosphokinase", "CPK by Outcome", "CPK (mcg/L)")

# Serum sodium plots
p5 <- hist_density_plot("serum_sodium", "Serum Sodium Distribution", "#16a085")
p6 <- boxplot_by_outcome("serum_sodium", "Serum Sodium by Outcome", "Serum Sodium (mEq/L)")

# Combine plots
(p1 + p2) / (p3 + p4) / (p5 + p6)
```
### Explore categorical variables
```{r,echo=FALSE, warning=FALSE, fig.width=10, fig.height=8, out.width="100%"}
# Function to create enhanced bar charts for binary variables
plot_binary_variable <- function(variable, var_label, color1, color2) {
  # Prepare data
  counts_data <- as.data.frame(table(data[[variable]], data$DEATH_EVENT))
  names(counts_data) <- c("Variable", "Outcome", "Count")
  counts_data$Variable <- ifelse(counts_data$Variable == 0, "No", "Yes")
  counts_data$Outcome <- ifelse(counts_data$Outcome == 0, "Survived", "Died")
  
  # Create plot
  p <- ggplot(counts_data, aes(x = Variable, y = Count, fill = Outcome)) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7) +
    scale_fill_manual(values = c(color1, color2)) +
    geom_text(aes(label = Count), 
              position = position_dodge(width = 0.7), 
              vjust = -0.5, size = 3) +
    labs(title = var_label,
         x = NULL,
         y = "Count") +
    my_theme
  return(p)
}

# Create plots for all binary variables
p1 <- plot_binary_variable("anaemia", "Anaemia", "#3498db", "#e74c3c")
p2 <- plot_binary_variable("diabetes", "Diabetes", "#3498db", "#e74c3c")
p3 <- plot_binary_variable("high_blood_pressure", "High Blood Pressure", "#3498db", "#e74c3c")
p4 <- plot_binary_variable("sex", "Sex (1=Male, 0=Female)", "#3498db", "#e74c3c")
p5 <- plot_binary_variable("smoking", "Smoking", "#3498db", "#e74c3c")

# Arrange plots in a grid
(p1 + p2) / (p3 + p4) / (p5 + plot_spacer()) 
```
### Correlation analysis

```{r, fig.width=10, fig.height=8,echo=FALSE, warning=FALSE}
library(ggplot2)
library(reshape2)
library(viridis)

# Calculate correlation matrix
cor_matrix <- cor(data)

# Format the correlation values to display
cor_matrix_text <- round(cor_matrix, 3)

# Melt the correlation matrix for ggplot
melted_cor <- melt(cor_matrix)
colnames(melted_cor) <- c("Var1", "Var2", "value")

# Add text labels from the rounded matrix
melted_cor$text_label <- melt(cor_matrix_text)$value

# Create the correlation heatmap
ggplot(melted_cor, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = text_label), color = "black", size = 3) +
  scale_fill_gradient2(
    low = "#4575b4",    # Blue for negative correlations
    mid = "white",      # White for zero
    high = "#d73027",   # Red for positive correlations
    midpoint = 0,
    limit = c(-1, 1),
    name = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14)
  ) +
  coord_fixed() +
  labs(title = "The correlation among features")
```
\section{Exploratory Data Analysis Summary}

This exploratory data analysis examines a dataset of 299 patients who had heart failure, focusing on patient characteristics and mortality outcomes.

\subsection{Dataset Overview}
The dataset consists of 13 clinical features collected during patient follow-up periods:
\begin{itemize}
  \item Demographic: age (years), sex (binary: 1=male, 0=female)
  \item Clinical conditions: anaemia, diabetes, high blood pressure, smoking (all binary: 1=yes, 0=no)
  \item Heart measurements: ejection fraction (percentage), creatinine phosphokinase (CPK, mcg/L)
  \item Blood measurements: platelets (kiloplatelets/mL), serum creatinine (mg/dL), serum sodium (mEq/L)
  \item Outcome: time (follow-up period in days), death event (1=yes, 0=no)
\end{itemize}

\subsection{Target Variable Distribution}
The dataset shows an imbalanced distribution of the target variable (death event):
\begin{itemize}
  \item 203 patients survived (67.9\%)
  \item 96 patients died (32.1\%)
\end{itemize}

\subsection{Numerical Variables}
Analysis of key numerical variables reveals several patterns:

\textbf{Age:} Patients' ages range from approximately 40 to 95 years, with a roughly normal distribution. Deceased patients tend to be slightly older than survivors, though there is substantial overlap.

\textbf{Ejection Fraction:} Shows a multimodal distribution with peaks around 20\%, 38\%, and 60\%. Lower ejection fraction values appear associated with higher mortality, supporting its known clinical significance in heart failure.

\textbf{Serum Creatinine:} Exhibits a right-skewed distribution with most values below 2.5 mg/dL. Deceased patients show slightly higher creatinine levels, indicating possible renal dysfunction association with mortality.

\textbf{Platelets:} Normally distributed around 250,000 kiloplatelets/mL, with no clear difference between survivors and non-survivors.

\textbf{CPK:} Highly right-skewed with most values below 2000 mcg/L but some extreme outliers exceeding 7000. No substantial difference observed between outcome groups.

\textbf{Serum Sodium:} Normally distributed around 135-140 mEq/L, with deceased patients showing slightly lower values, suggesting hyponatremia might be associated with poorer outcomes.

\subsection{Categorical Variables}
The categorical variable analysis shows several patterns:

\textbf{Anaemia:} Present in 129 patients (43\%), with higher mortality proportion among anaemic patients.

\textbf{Diabetes:} Present in 125 patients (42\%), with slightly higher mortality proportion.

\textbf{High Blood Pressure:} Present in 105 patients (35\%), with slightly higher mortality rate among those without hypertension.

\textbf{Sex:} The dataset includes 194 males (65\%) and 105 females. Males show a slightly higher mortality rate.

\textbf{Smoking:} Present in 96 patients (32\%), with smokers showing lower mortality compared to non-smokers.

\subsection{Correlation Analysis}
The correlation heatmap reveals several significant relationships:

\textbf{Strong correlations:}
\begin{itemize}
  \item Death event and time (-0.53): Negative correlation indicating patients who died had shorter follow-up periods
  \item Sex and smoking (0.45): Males more likely to be smokers
  \item Death event and ejection fraction (-0.27): Lower ejection fraction associated with higher mortality
  \item Death event and serum creatinine (0.29): Higher serum creatinine associated with higher mortality
  \item Age and death event (0.25): Older age associated with higher mortality
\end{itemize}

\textbf{Moderate correlations:}
\begin{itemize}
  \item High blood pressure and time (-0.20): Patients with hypertension had shorter follow-up periods
  \item Serum sodium and death event (-0.20): Lower sodium levels associated with mortality
  \item Anaemia and creatinine phosphokinase (-0.19): Inverse relationship
\end{itemize}

\subsection{Key Findings}
\begin{enumerate}
  \item The three strongest predictors of mortality appear to be time (follow-up period), ejection fraction, and serum creatinine.
  \item Age shows a positive correlation with mortality risk, confirming the expected clinical relationship.
  \item Ejection fraction is inversely related to mortality, consistent with heart failure pathophysiology.
  \item Serum abnormalities (creatinine, sodium) show associations with mortality, suggesting the importance of metabolic factors.
  \item Categorical variables (anaemia, diabetes, hypertension, sex, smoking) show some patterns with mortality but with weaker associations than continuous measurements.
\end{enumerate}

These findings align with clinical understanding of heart failure risk factors and highlight the complex interplay between demographic, cardiac, and metabolic parameters in predicting mortality in heart failure patients.


