=--
title: "R Code for Survival Analysis Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the library "survival" which includes coxph() 

Load a library for parametric regression; these are rarely used because Cox's semiparametric model is favored. 

Example that contrasts (finds similarities) between CoxPH and Weibull using the pbc data
Note: Weibull is the only model which is both a Proportional Hazards model and accelerated failure time model (i.e. log-linear model for time)

```{r}
library(survival)
#install.packages("SurvRegCensCov")
library(SurvRegCensCov)
help(pbc)
names(pbc)
dead.tx <- pbc[,"status"] >= 1
```

# Univariable time-to-event models for hazard of death or transplant

```{r}
WeibullReg(Surv(time, dead.tx) ~ trt, data=pbc)
summary(coxph(Surv(time, dead.tx) ~ trt, data=pbc))

```

Multivariable Weibull and Cox models
Comment on the similarities and/or differences
```{r}
WeibullReg(Surv(time, dead.tx) ~ albumin + bili + chol + protime, data=pbc)

summary(coxph(Surv(time, dead.tx) ~ albumin + bili + chol + protime, data=pbc))

```


Using pbc again, even more multivariable
Including several covariates at once using a matrix
As well as separately listed covariates
```{r}
Covar <- as.matrix(pbc[,7:20])
summary(os <- coxph(Surv(time, dead.tx) ~ age + sex + Covar, data=pbc))
#WeibullReg(Surv(time, dead.tx) ~ age + sex + Covar, data=pbc)

```

LASSO version of Cox
LASSO is often used for model selection because the least important variables are assigned zero coefficients
```{R}
library(glmnet)
female = pbc$sex == "f"
X = cbind(age=pbc$age, female, Covar)
Miss <- is.na(pbc$time) | is.na(dead.tx) | rowSums(is.na(X)) > 0
o.cv <- cv.glmnet(y = Surv(pbc$time, dead.tx)[!Miss], x=X[!Miss, ], famil="cox")
plot(o.cv)

o <- glmnet(y = Surv(pbc$time, dead.tx)[!Miss], x=X[!Miss, ], family="cox")
plot(o)

# Coefficients at best penalty according to cross validated partial likelihood
round(o$beta[,o$lambda==o.cv$lambda.min],4)
# those that are nonzero
dimnames(X)[[2]][o$beta[,o$lambda==o.cv$lambda.min] > 0]
```


Rescaling time axis makes no difference to coefficients and hazard ratios
```{r}
round(coef.1 <- summary(os <- coxph(Surv(time, dead.tx) ~ age + sex + Covar, data=pbc))$coef,3)

Years <- pbc$time/365.25
round(coef.2 <- summary(os <- coxph(Surv(Years, dead.tx) ~ age + sex + Covar, data=pbc))$coef,3)
# Are both sets of coefficients the same?
coef.1 == coef.2
```


Investigating Proportionality of Hazards for treatment effect
```{r}
# Look for proportional cumulative hazards
plot(survfit(Surv(time, dead.tx) ~ trt, data=pbc), fun="cumhaz", lwd=3, col=2:3, mark.time=FALSE, main="Cumulative Hazard Estimate", xlab="Days")

# Look for parallel log cumulative hazards
plot(survfit(Surv(time, dead.tx) ~ trt, data=pbc), fun="cloglog", lwd=3, col=2:3, mark.time=FALSE, main="Log Cumulative Hazard", xlab="Days")
```

Investigating Proportionality of Hazards for the continuous variable, prothrombin time
```{r}
protime.3 <- cut(pbc[,"protime"], c(0,10,11,Inf))

plot(survfit(Surv(time, dead.tx) ~ protime.3, data=pbc), fun="cloglog", lwd=3, col=2:4, mark.time=FALSE, main="Log Cumulative Hazard", xlab="Days")

# Schoenfeld Residual test of proportionality
summary(os <- coxph(Surv(time, dead.tx) ~ protime, data=pbc))
(o.zph <- cox.zph(os))
plot(o.zph$time, o.zph$y)
```

# Can test several at once
```{r}
summary(os <- coxph(Surv(time, dead.tx) ~ age + sex + protime, data=pbc))
cox.zph(os)

```

Hazard Ratios stratified by time
```{r}
quantile(pbc[dead.tx==1,"time"], c(0.33,0.67))
dead.tx.1 <- ifelse(pbc[,"time"]<800, dead.tx, 0)
dead.tx.2 <- ifelse(pbc[,"time"]>=800 & pbc[,"time"]<1600, dead.tx, 0)
dead.tx.3 <- ifelse(pbc[,"time"]>=1600, dead.tx, 0)
summary(os1 <- coxph(Surv(time, dead.tx.1) ~ protime, data=pbc))$coef
summary(os2 <- coxph(Surv(time, dead.tx.2) ~ protime, data=pbc))$coef
summary(os3 <- coxph(Surv(time, dead.tx.3) ~ protime, data=pbc))$coef

```


Testing a categorical or ordinal variable, stage
```{r}
library(lmtest)
summary(os1 <- coxph(Surv(time, dead.tx) ~ trt + age + sex, data=pbc))
summary(os2 <- coxph(Surv(time, dead.tx) ~ trt + age + sex + stage, data=pbc))
fstage <- factor(pbc$stage)
summary(os3 <- coxph(Surv(time, dead.tx) ~ trt + age + sex + fstage, data=pbc))
```

# likelihood ratio test
```{r}
lrtest(os1, os3)
# or
anova(os3)

```


# Stratified baseline hzards: An interesting other way to stratify with Cox's model
This approach avoids parameterization of the effect of the strata
```{r}
summary(os3 <- coxph(Surv(time, dead.tx) ~ trt + age + stage + sex, data=pbc))
summary(os3 <- coxph(Surv(time, dead.tx) ~ trt + age + stage + strata(sex), data=pbc))

summary(os3 <- coxph(Surv(time, dead.tx) ~ trt + age + sex + sex + strata(stage), data=pbc))

```
# Illustration of start stop
```{r}
?cgd
table(cgd$id)
plot(cgd$tstop, cgd$id, xlab=c(0,max(cgd$tstop)), pch=16)
points(cgd$tstart, cgd$id, pch=16)
for (i in unique(cgd$id)) {
  df <- cgd[cgd$id==i,]
  for (j in 1:dim(df)[1]) lines(c(df$tstart[j], df$tstop[j]), rep(i,2))
}

table(cgd$treat)
coxph(Surv(tstart,tstop,status)~treat, data=cgd)

```
# Illustration of frailties: NOT for use for id
```{r}
coxph(Surv(tstart,tstop,status)~treat + frailty(center), data=cgd)
coxph(Surv(tstart,tstop,status)~treat + cluster(center), data=cgd)

```



Investigating departures from (log)linear effects of continuous effects
```{r}
# is the effect of age on the hazard log-linear ?
summary(osa <- coxph(Surv(time, dead.tx) ~ pspline(age, df=3), data=pbc))
plot(pbc$age, predict(osa))

# Can be multivariable
summary(osa <- coxph(Surv(time, dead.tx) ~ pspline(age, df=3) + sex + protime, data=pbc))

# is the effect of protime on the hazard log-linear ?
summary(osp <- coxph(Surv(time, dead.tx) ~ pspline(protime, df=3) + sex, data=pbc))


```


Example of start-stop data, aka "counting process"
```{r}
help(heart)
View(heart)
(Surv.heart <- with(heart, Surv(time=start, time2=stop, event=event, type="counting")))
summary(Surv.heart)
summary(coxph(Surv.heart ~ age + year + transplant, data=heart))
```




Illustration that Hazards Cross Before the Survival Curves Cross
```{r}
par(mfcol=c(2,2))
Time <- seq(from=0,to=(ULT<-3),length=1000)[-1]
rho <- 0.25
Haz0 <- Time^rho
Haz1 <- Time^(-rho)
plot(Time, Haz0, xlab="Time", main="Hazard", lwd=3, col=2, type="l", ylim=range(c(Haz0, Haz1)))
lines(Time, Haz1, col=5)
abline(v=1, lty=3)
plot(Time, exp(-Time^(rho+1)/(rho+1)), xlab="Time", main="Survival", lwd=3, col=2, type="l")
lines(Time, exp(-Time^(-rho+1)/(-rho+1)) , col=5)
abline(v=1, lty=3)

Haz0 <- Time^rho
Haz1 <- (1/2)*Time^(-rho)
plot(Time, Haz0, xlab="Time", main="Hazard", lwd=3, col=2, type="l")
lines(Time, Haz1, col=5)
abline(v=(1/2)^(1/(2*rho)), lty=3)
plot(Time, exp(-Time^(rho+1)/(rho+1)), xlab="Time", main="Survival", lwd=3, col=2, type="l")
lines(Time, exp(-(1/2)*Time^(-rho+1)/(-rho+1)) , col=5)
abline(v=(1/2)^(1/(2*rho)), lty=3)
```


Hazard Ratio stratified by time
Simulation of data with crossing hazards
```{r}
n <- 5000
T0 <- rexp(n) / (BaselineHaz0<-0.5)
CumHaz1 <- function(Time) {
  BaselineHaz0 * (3*Time - 2.99*log(1+Time))
  }
inverse.f <- function(y, f) {
  g <- function(x) {f(x)-y}
  uniroot(g, lower=0, upper=10^8)
  }
T1 <- rep(NA, n)
for (i in 1:n) 
  T1[i] <- inverse.f(y=-log(runif(1)), f=CumHaz1)$root
X <- runif(n) < 0.5
Time <- ifelse(X, T1, T0)
Censor <- runif(n)
Censor <- Censor * quantile(Time/Censor, prop.events<-0.50)
FUT <- pmin(Time, Censor)
Event <- ifelse(Time < Censor, 1, 0)

par(mfrow=c(1,1))
plot(survfit(Surv(FUT, Event) ~ X), fun="event", lwd=3, col=2:3, ylab="Mortality", mark.time=FALSE, xlab="Time")

X <- as.numeric(X)
summary(coxph(Surv(FUT, Event) ~ X, robust=TRUE))$coef
# HR Stratifed by followup time cut at 1 and 2
summary(coxph(Surv(FUT, ifelse(FUT<1, Event, 0)) ~ X))$coef
summary(coxph(Surv(FUT, ifelse(FUT>=1 & FUT<2, Event,0)) ~ X))$coef
summary(coxph(Surv(FUT, ifelse(FUT>=2, Event, 0)) ~ X))$coef
```




# Illustration of Clustered (aka, MultiLevel) Data using simulated data
```{r}
n.cluster <- 100
Cluster <- 1:n.cluster
Frailty <- rnorm(n.cluster)
n.per.cluster <- rpois(n.cluster, lambda=10)
N <- sum(n.per.cluster)
Cluster.ID <- rep(Cluster, times=n.per.cluster)
Long.Frailty <- rep(Frailty, times=n.per.cluster)
X1 <- rnorm(N)
X2 <- rnorm(N)
X3 <- rnorm(N)
Hazard <- exp(1*X1 + 0*X2 + 0*X3 + 1.0*Long.Frailty)
Time <- rexp(N) / Hazard
Censor <- rexp(N)
FUT <- pmin(Time, Censor)
Event <- ifelse(Time < Censor, 1, 0)

coxph(Surv(FUT, Event) ~ X1 + X2 + X3)
coxph(Surv(FUT, Event) ~ X1 + X2 + X3 + frailty(Cluster.ID))
coxph(Surv(FUT, Event) ~ X1 + X2 + X3 + cluster(Cluster.ID))
```



Illustration using simulated data
```{r}
n<-10000
Z1 <- rnorm(n)
Z2 <- rnorm(n)
Z3 <- rnorm(n)
X <- 0.5*Z1 + 0.2*Z2 + 0.0*Z3 + rnorm(n)
HR <- exp(log(2)*X + -0.5*Z1 + 0.5*Z2 + 1*Z3)
Time <- (rexp(n) / HR)^(1/3)
Censor <- runif(n)
Censor <- Censor * quantile(Time/Censor, q<-0.5)
FUT <- pmin(Time, Censor)
Event <- ifelse(Time < Censor, 1, 0)

WeibullReg(Surv(FUT, Event) ~ X)
coxph(Surv(FUT, Event) ~ X)

WeibullReg(Surv(FUT, Event) ~ X+Z1+Z2+Z3)
summary(coxph(Surv(FUT, Event) ~ X+Z1+Z2+Z3))

```


# Noncollapsibility 
```{r}
summary(coxph(Surv(FUT, Event) ~ X+Z1+Z2))$coef
summary(coxph(Surv(FUT, Event) ~ X+Z1+Z2+Z3))$coef
```






# Competing risks: e.g. death and transplant compete with each other
```{r}
#install.packages("cmprsk")
library(cmprsk)
(ocr <- cuminc(pbc$time/365.25, pbc$status))
plot(ocr)

(okm <- survfit(Surv(pbc$time/365.25, pbc$status>0)~1))
lines(okm$time, 1 - okm$surv, col=2, type="s")

?crr


```
